### 2017-06-07 src/3_TimeSeries.tex
Chapter 4: Springer paper finished

\section{Vector error correction model}\label{sec:vecm}
Vector error correction model (VECM)  \cite{engle87} describe the joint behaviour of a set of cointegrated I(1) time series \cite{banerjee1993} and can be derived from the simple VAR model.

@@

Transposing each equation of the system (\ref{eq:vecfull}) we can write
the VECM($p$) model in block-matrix form as:
\begin{equation}\label{eq:simpleveceq}
\mathbf{Y} = 
\mathbf{A} \mathbf{X} + 
\mathbf{E} \, , 
\end{equation}

@@

\begin{align}
\label{eq:vecY}
\mathbf{Y}
= \begin{bmatrix}
   \Delta\mathbf{y}_{p+1}^\top \\
   \Delta\mathbf{y}_{p+2}^\top \\
   \vdots \\
   \Delta\mathbf{y}_N^\top
   \end{bmatrix}
\end{align}

>>>>

\section{Vector error correction model}\label{sec:vecm}
The VAR model cannot be used with non-stationary variables. VECM ,
developed by \cite{engle87}, is also a linear model for I(1) variables that are
also cointegrated, see \cite{banerjee1993}. If cointegration exists, variable
differences are stationary and they introduce an error correction term which
adjusts coefficients to bring the variables back to equilibrium.

@@

Transposing each equation of the system (\ref{eq:vecfull}) we can write
the VECM($p$) model in block-matrix form as:
\begin{equation}\label{eq:simpleveceq} %vareq}
\mathbf{B} = 
\mathbf{A} \mathbf{X} + 
\mathbf{E} \, , 
\end{equation}

@@

\begin{align}
\label{eq:vecY}
\mathbf{B}
= \begin{bmatrix}
   \Delta\mathbf{y}_{p+1}^\top \\
   \Delta\mathbf{y}_{p+2}^\top \\
   \vdots \\
   \Delta\mathbf{y}_N^\top
   \end{bmatrix}
\end{align}

### 2017-06-07 src/4_OnlineLearning.tex
Chapter 4: Springer paper finished

Forecast performance is evaluated using different methods. We have chosen three
measures usually used:
\begin{description}
\item[MAPE] Mean Average Percent Error which presents forecast errors as a
percentage.
\begin{equation}\label{eq:MAPE}
\text{MAPE} = \frac{1}{N} \sum_{t=1}^{N} 
\frac{\left|\mathbf{y}_t-\hat{\mathbf{y}}_t\right|}{\left|\mathbf{y}_t\right|}
 \times 100 
\end{equation}
\item[MAE] Mean Average Error which measures the distance between forecasts to the
true value.
\begin{equation}\label{eq:MAE}
\text{MAE} = \frac{1}{N} \sum_{t=1}^{N} 
\left| 
\mathbf{y}_t-\hat{\mathbf{y}}_t
\right| 
\end{equation}
\item[RMSE] Root Mean Square Error also measures the distance between forecasts
to the true values but, unlike MAE, large deviations from the true value have a
large impact on RMSE due to squaring forecast error.
\begin{equation}\label{eq:RMSE}
\text{RMSE} = \sqrt{
\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}}
\end{equation}
\end{description}

>>>>

Forecast performance is evaluated using different methods. We have chosen three
measures usually used:
\begin{description}
\item[MAPE] Mean Average Percent Error which presents forecast errors as a
percentage.
\begin{equation}\label{eq:MAPE}
\text{MAPE} = \frac{1}{N} \sum_{t=1}^{N} 
\frac{\left|\mathbf{y}_t-\hat{\mathbf{y}}_t\right|}{\left|\mathbf{y}_t\right|}
 \times 100 
\end{equation}
\item[MAE] Mean Average Error which measures the distance between forecasts to the
true value.
\begin{equation}\label{eq:MAE}
\text{MAE} = \frac{1}{N} \sum_{t=1}^{N} 
\left| 
\mathbf{y}_t-\hat{\mathbf{y}}_t
\right| 
\end{equation}
\item[MSE]  Mean Square Error measures the distance between forecasts
and the true values and large deviations from the true value have a
large impact due to squaring forecast error.
\begin{equation}\label{eq:MSE}
\text{MSE} = 
\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}
\end{equation}
\item[RMSE] Root Mean Square Error also measures the distance between forecasts
to the true values but, unlike MAE, large deviations from the true value have a
large impact on RMSE due to squaring forecast error.
\begin{equation}\label{eq:RMSE}
\text{RMSE} = \sqrt{
\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}}
\end{equation}
\item[$U$-statistic] the Theil's $U$-statistic, presented by
\cite{theil1966}, is a unit free measure obtained as the ratio between the root
MSE (RMSE) of a model and the RMSE of the naive random walk model. A
$U$-statistic less than 1 implies the performance is better than the naive
model.
\end{description}

### 2017-06-07 src/5_1_Proposal.tex
my citation
Chapter 4: Springer paper finished

\chapter{Fast and adaptive cointegration based model
for forecasting high frequency financial time series}

@@

Cointegration relations change in time and therefore VECM parameters must be
updated when new data is available. We studied how forecasting performance is
affected when VECM parameters and the length of historical data used change in
time. We observed that the number of cointegration relationships varies with the
length of historical data used. Moreover, parameters that increased these
relationships in time led to better forecasting performance. Our proposal,
called an Adaptive VECM (AVECM) is to make a parameters grid search that
maximises the number of cointegration relationships in the near past. To ensure
the search can be executed fast enough, we used a distributed environment 
\cite{Arce2017}.

@@

@@

%\section{Background}
%\label{sec:background}
%
%\subsection{Integration and Cointegration}\label{sec:coint}\  
%Following \cite{johansen1995} we shall say that a stochastic process
%$Y_t$ which satisfies $Y_t-E(Y_t) = \sum_{i=0}^\infty C_i\,\varepsilon_{t-i}$ is
%called $I(0)$, and then we shall write $Y_t\sim I(0)$, whenever
%$\sum_{i=0}^\infty C_i \neq 0$ and $\sum_{i=0}^\infty C_i\,z^i$ converges for
%$z\in\mathbb{C}$ with $|z|<1$.  It is understood that the condition
%$\varepsilon_t\sim iid(0,\sigma^2)$ holds.
%
%A (vector) time series $\mathbf{y}_t$ is said to be {\em integrated of order\/}
%$d$, and then we shall write $\mathbf{y}_t\sim I(d)$, whenever after $d$ times
%(discrete) differentiation an stationary process is
%obtained, see \cite{banerjee1993}; more precisely, whenever
%$(1-L)^d\,\mathbf{y}_t\sim\text{I(0)}$, where $L$ is the usual lag operator:
%$(1-L)\,\mathbf{y}_t = \Delta\mathbf{y}_t = \mathbf{y}_t-\mathbf{y}_{t-1}$ for
%all $t$.  
%
%Note that this definition includes the scalar case as time series of vectors of
%dimension 1; in this scalar case we will write the time series in nonbold
%format.
%
%Let $\mathbf{y}_t^\nu$, $\nu=1,\dots,l$, be a set of $l$ vector time series of
%order $I(1)$.  They are said to be {\em cointegrated\/} if a vector
%$\beta=[\beta(1),\dots,\beta(l)]^\top \in \mathbb{R}^p$ exists, such that the
%time series,
%\begin{equation}
%\mathbf{Z}_t:= 
%\sum_{\nu=1}^l \beta(\nu)\,\mathbf{y}_t^\nu\,\sim\,\text{I(0)}\,.
%\end{equation}
%In other words, a set of $I(1)$ time series is said to be cointegrated if a
%linear combination of them exists, which is I(0).
%
%
%\subsection{Vector Autorregresive Models}\label{sec:varvec}
%
%Vector error correction model (VECM) describe the joint behaviour of a set of
%variables and can be derived from the simple Vector Autoregressive model (VAR)
%presented by \cite{sims1980}.  The VAR($p$) model is a framework describing the
%behaviour of a set of $l$ endogenous and stationary variables as a linear
%combination of their last $p$ values, where $l,p\in\mathbb{N}$.  In our case,
%each one of these $l$ variables is a scalar time series $y_{\lambda,t}$,
%$\lambda=1,\dots,l$, and we represent them all together at time $t$ by the
%vector time series:
%\begin{equation}
%\label{eq:variables}
%\mathbf{y}_t = 
%\begin{bmatrix} y_{1,t} & y_{2,t} & \dots & y_{l,t} \end{bmatrix}^\top.
%\end{equation}
%\noindent
%Notice that the vectors $\mathbf{y}_t$ are assumed to be $l$-dimensional.
%
%The VAR($p$) model describes the behaviour of a dependent variable in terms of
%its own lagged values and the lags of the others variables in the system. The
%model with $p$ lags is formulated as the system:
%\begin{align}
%\label{eq:var}
%\mathbf{y}_t 
%= \boldsymbol{\Phi}_1 \mathbf{y}_{t-1} +
%  \boldsymbol{\Phi}_2 \mathbf{y}_{t-2} + \dots +
%  \boldsymbol{\Phi}_p\mathbf{y}_{t-p} +
%  \mathbf{c} + \boldsymbol{\epsilon}_t \nonumber \\
%t=p+1,\dots,N,
%\end{align}
%\noindent where 
%$\boldsymbol{\Phi}_1, \boldsymbol{\Phi}_2,\dots,\boldsymbol{\Phi}_p$
%are $l\times l$-matrices of real coefficients,
%$\boldsymbol{\epsilon}_{p+1},
% \boldsymbol{\epsilon}_{p+2}, \dots, \boldsymbol{\epsilon}_N$ 
%are error terms, $\mathbf{c}$ is a constant vector and $N$ is the total
%number of samples.
%
%Notice that, regarding our notation of section (\ref{sec:coint}),
%we have here 
%$\mathbf{y}_t^0 = \mathbf{y}_t$,
%$\mathbf{y}_t^\nu = \mathbf{y}_{t-\nu}$ and
%the $\lambda$-th component of the vector time series $\mathbf{y}_t^\nu$
%is the scalar time series $y_{\lambda,t}^\nu$, where $\nu=1,\dots,p$ and
%$\lambda=1,\dots,l$.
%
%However, the VAR model cannot be used with non-stationary variables. VECM ,
%developed by \cite{engle87}, is also a linear model for I(1) variables that are
%also cointegrated, see \cite{banerjee1993}. If cointegration exists, variable
%differences are stationary and they introduce an error correction term which
%adjusts coefficients to bring the variables back to equilibrium. 
%
%
%It is obtained re-writing equation (\ref{eq:var}) in terms of the new
%variable $\Delta\mathbf{y}_t=\mathbf{y}_t-\mathbf{y}_{t-1}$.
%The VECM model, expressed in terms those differences, takes the form:
%\begin{equation}\label{eq:vec}
%\Delta \mathbf{y}_t 
%= \boldsymbol{\Omega}\,\mathbf{y}_{t-1}
%  + \sum_{i=1}^{p-1} \boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
%  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
%\end{equation}
%\noindent
%where the coefficients matrices $\boldsymbol{\Phi}_i^*$ and 
%$\boldsymbol{\Omega}$, expressed in terms of the matrices
%$\boldsymbol{\Phi}_i$ of (\ref{eq:var}), are:
%\begin{align*}
%\boldsymbol{\Phi}_i^* 
%&:= -\sum_{j=i+1}^{p}\boldsymbol{\Phi}_j\,, \\
%\boldsymbol{\Omega}
%&:= -\left( \mathbb{I} - \boldsymbol{\Phi}_1 - \dots 
%    - \boldsymbol{\phi}_p \right)\,. 
%\end{align*}
%The following well known properties of the matrix $\boldsymbol{\Omega}$, see
%\cite{johansen1995}, will be useful in the sequel:
%\begin{itemize}
%\item
%If $\boldsymbol{\Omega} = \mathbf{0}$, there is no cointegration.
%\item 
%If $rank(\boldsymbol{\Omega})=l$, i.e., if $\boldsymbol{\Omega}$ has
%full rank, then the time series are not I(1) but stationary.
%\item
%If $rank(\boldsymbol{\Omega})=r$, $0<r<l$, then there is cointegration
%and the matrix $\boldsymbol{\Omega}$ can be expressed as
%$\boldsymbol{\Omega}=\boldsymbol{\alpha\beta}^\top$, where $\boldsymbol{\alpha}$
%and $\boldsymbol{\beta}$ are
%$l\times r$ matrices and
%$\text{rank}(\boldsymbol{\alpha})=\text{rank}(\boldsymbol{\beta})=r$.
%\item
%The columns of $\boldsymbol{\beta}$ contains the cointegration vectors and the rows of
%$\boldsymbol{\alpha}$ correspond with the adjusted vectors. 
%$\boldsymbol{\beta}$ is obtained by Johansen procedure,see \cite{johansen1988},
%whereas $\boldsymbol{\alpha}$ has to be determined as a variable in the VECM.
%\end{itemize}
%It is worth noticing that the factorization of the matrix
%$\boldsymbol\Omega$ is not unique, since for any $r \times r$
%non-singular matrix $\mathbf{H}$, $\boldsymbol{\alpha}^*:=\boldsymbol{\alpha}\mathbf{H}$,
%and $\boldsymbol{\beta}^*=\boldsymbol{\beta}(\mathbf{H}^{-1})^\top$ we have
%$\boldsymbol{\alpha\beta}^\top=\boldsymbol{\alpha}^*(\boldsymbol{\beta}^*)^\top$.
%If cointegration exists, then equation (\ref{eq:vec}) can be written
%as follows:
%\begin{equation}\label{eq:vecfull}
%\Delta\mathbf{y}_t 
%= \boldsymbol{\alpha\beta}^\top\mathbf{y}_{t-1} 
%  + \sum_{i=1}^{p-1}\boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
%  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
%\end{equation}
%\noindent
%which is a VAR model but for time series differences.
%
%
%Transposing each equation of the system (\ref{eq:vecfull}) we can write
%the VECM($p$) model in block-matrix form as:
%\begin{equation}\label{eq:vareq}
%\mathbf{B} = 
%\mathbf{A} \mathbf{X} + 
%\mathbf{E} \, , 
%\end{equation}
%%
%\noindent where $\mathbf{B}$ dimension is $((N-p)\times l)$, $\mathbf{A}$
%dimension is $((N-p)\times(r+(p-1)l +1))$, $\mathbf{X}$ dimension is $((r+(p-1)l
%+1)\times l)$ and $\mathbf{E}$ dimension is $((N-p)\times l)$:
%%
%\begin{alignat}{3}
%\mathbf{B}
%&= \begin{bmatrix}
%   \Delta\mathbf{y}_{p+1}^\top \\
%   \Delta\mathbf{y}_{p+2}^\top \\
%   \vdots \\
%   \Delta\mathbf{y}_N^\top
%   \end{bmatrix}
%&\quad
%\mathbf{X}
%&= \begin{bmatrix}
%   \boldsymbol{\alpha}^\top \\
%   \boldsymbol{\Phi}_1^{*\top} \\
%   \boldsymbol{\Phi}_2^{*\top} \\
%   \vdots \\
%   \boldsymbol{\Phi}_{p-1}^{*\top} \\
%   \mathbf{c}^\top
%   \end{bmatrix}
%&\quad
%\mathbf{E}
%&= \begin{bmatrix}
%   \boldsymbol{\epsilon}_{p+1}^\top \\
%   \boldsymbol{\epsilon}_{p+2}^\top \\
%   \vdots \\
%   \boldsymbol{\epsilon}_N^\top \\
%   \end{bmatrix}
%\end{alignat}
%\noindent and 
%\begin{align}
%\mathbf{A} 
%&= \begin{pmat}[{....|}]
%   \mathbf{y}_p^\top \boldsymbol{\beta} & \Delta \mathbf{y}_p^\top & \Delta\mathbf{y}_{p-1}^\top & \dots 
%                    & \Delta\mathbf{y}_2^\top & 1 \cr
%   \mathbf{y}_{p+1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{p+1}^\top & \Delta\mathbf{y}_p^\top & \dots
%                       & \Delta\mathbf{y}_3^\top & 1 \cr
%   \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \cr
%   \mathbf{y}_{N-1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{N-1}^\top & \Delta\mathbf{y}_{N-2}^\top & \dots 
%                       & \Delta\mathbf{y}_{N-p-1}^\top & 1 \cr
%   \end{pmat}\, .
%\label{eq:Amatrix}
%\end{align}
%Taking into account the error term $\mathbf{E}$, equation~(\ref{eq:vareq}) 
%can be solved with respect to $\mathbf{X}$ using the OLS estimation.

\section{Methodology} \label{sec:51methodology}
\subsection{Motivation}

@@

\subsection{Evaluation methods}

Forecast performance was evaluated using two different methods which are
frequently used in finance:
\begin{description}
\item[MSE]  Mean Square Error measures the distance between forecasts
and the true values and large deviations from the true value have a
large impact due to squaring forecast error.
\begin{equation}\label{eq:MSE}
\text{MSE} = 
\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}
\end{equation}
\item[$U$-statistic] the Theil's $U$-statistic, presented by
\cite{theil1966}, is a unit free measure obtained as the ratio between the root
MSE (RMSE) of a model and the RMSE of the naive random walk model. A
$U$-statistic less than 1 implies the performance is better than the naive
model.
\end{description}

@@

Algorithms AVECM, ARIMA and the naive random walk were tested using four days of
data (from the 12th to the 15th of August 2014).  For AVECM we considered
different number of iterations (parameter $m$ in algorithm \ref{alg:AVECM}): 10,
50 and 100. We tried 12, 24 and 47 different pair of combinations for $L$ and
$p$. Possible values for $L$ were in the interval $[2,14]$ hours and $p$ can
have values in between $[1,5]$. Best AVECM performance was compared against
ARIMA and the random walk model.  Table~\ref{tab:stats} shows the out-of-sample
performance measures: MSE and $U$-statistic for AVECM and ARIMA. In terms of
both measures we found that AVECM is superior to ARIMA and the naive random walk
model. We also included the p-value that proves that the difference in the MSE
is significant at the 99\% significance level in three of the four currency
rates and at 90\% in the case of GBPUSD. The $U$-statistic shows that AVECM and
ARIMA are superior to the naive random walk model and that our proposal is also
superior to ARIMA.

\begin{table}[!htbp]
\caption{AVECM performance}
\label{tab:stats}
\begin{center}
\begin{tabular}{ccccccc}

@@

To determine $L$ and $p$ based on maximising the percentage of cointegration
requires use of the Johansen method which is a computationally expensive
routine. This procedure is done by the function
get\_best\_params($Ls,ps,m,\mathbf{Y}$) shown in algorithm \ref{alg:AVECM}.  In
order to improve the execution time of this search, our proposal included a
parallel search of VECM parameters using high performance computing.  The main
objective was to obtain a response before a new data arrived in the following 10
seconds.

@@

Execution time depends directly on $L$, $p$ and $nparams$ used, since they
determine the size of matrix $\mathbf{A}$ (see equation \ref{eq:Amatrix}) and
therefore affect the OLS function execution time.  Therefore, if we try more
combinations of $L$ and $p$ (increasing $nparams$) the serial algorithm will
take longer. For this financial time series we are interested in execution times
below 10 seconds  (the time series frequency).

>>>>

\chapter[Adaptive cointegration model for time series forecasting]{Fast and adaptive cointegration based model
for forecasting high frequency financial time series}

@@

Cointegration relations change in time and therefore VECM parameters must be
updated when new data is available. We studied how forecasting performance is
affected when VECM parameters and the length of historical data used change in
time. We observed that the number of cointegration relationships varies with the
length of historical data used. Moreover, parameters that increased these
relationships in time led to better forecasting performance. Our proposal,
called an Adaptive VECM (AVECM) is to make a parameters grid search that
maximises the number of cointegration relationships in the near past. To ensure
the search can be executed fast enough, we used a distributed environment .

@@

This work is published in \cite{Arce2017}.

@@

\section{Methodology} \label{sec:51methodology}

@@

@@

Algorithms AVECM, ARIMA and the naive random walk were tested using four days of
data (from the 12th to the 15th of August 2014).  For AVECM we considered
different number of iterations (parameter $m$ in algorithm \ref{alg:AVECM}): 10,
50 and 100. We tried 12, 24 and 47 different pair of combinations for $L$ and
$p$. Possible values for $L$ were in the interval $[2,14]$ hours and $p$ can
have values in between $[1,5]$. Best AVECM performance was compared against
ARIMA and the random walk model.  Table~\ref{tab:51stats} shows the out-of-sample
performance measures: MSE and $U$-statistic for AVECM and ARIMA. In terms of
both measures we found that AVECM is superior to ARIMA and the naive random walk
model. We also included the p-value that proves that the difference in the MSE
is significant at the 99\% significance level in three of the four currency
rates and at 90\% in the case of GBPUSD. The $U$-statistic shows that AVECM and
ARIMA are superior to the naive random walk model and that our proposal is also
superior to ARIMA.

\begin{table}[!htbp]
\caption{AVECM performance}
\label{tab:51stats}
\begin{center}
\begin{tabular}{ccccccc}

@@

To determine $L$ and $p$ based on maximising the percentage of cointegration
requires use of the Johansen method which is a computationally expensive
routine. This procedure is done by the function
get\_best\_params shown in algorithm \ref{alg:AVECM}.  In
order to improve the execution time of this search, our proposal included a
parallel search of VECM parameters using high performance computing.  The main
objective was to obtain a response before a new data arrived in the following 10
seconds.

@@

Execution time depends directly on $L$, $p$ and $nparams$ used, since they
determine the size of matrix $\mathbf{A}$ (see equation \ref{eq:vecA}) and
therefore affect the OLS function execution time.  Therefore, if we try more
combinations of $L$ and $p$ (increasing $nparams$) the serial algorithm will
take longer. For this financial time series we are interested in execution times
below 10 seconds  (the time series frequency).
